Potential of I/O Aware Workflows in Climate and Weather

### 2020-05-23 paper.tex
Some fixing

The vision proposed here is built on the belief that workflows composed of data, computing, and communication-intensive tasks should drive interfaces and hardware configurations to best support the programming models.
When delivered, this work will increase the opportunity for smarter scheduling of computing by considering storage in heterogeneous storage systems.
We illustrate the performance-impact on an example workload using a model built on measured performance data at DKRZ.

@@

\subsection{Data Center Infrastructure}

@@

The abstraction and automation of the I/O inside a workflow allow a runtime system to improve data placement and apply data reduction on heterogeneous storage systems.
Taking into consideration the architecture and workflow information, a smarter schedule can now be realised by exploiting the additional information.
Value and priority can influence fault-tolerance strategies and imply the quality of service for performance and availability.
Aspects like data reproducibility (can it be recomputed easily), type of the experiment (test, production), and runtime constraints for the overall and potential workflow could allow reducing costs and, hence, increasing scientific output.
Next, we outline the two core strategies and potential the proposed vision can bring to improve current workflows.
In the design proposed in this work (\cref{sec:design}), we will focus on the data placement strategy.

\paragraph{Strategy: Data Placement.}

@@

\paragraph{Strategy: Data Reduction.}

@@

ESDM is used as I/O middleware in the parallel application (with NetCDF or directly) and orchestrates the I/O according to a simplified ESDM configuration file (\texttt{esdm.conf}).
This file contains information about the available technology in the data centre, its I/O characteristics, and will be used to make decisions about how to prioritise I/O targets.
In the example presented in \Cref{lst:esdm.conf}, we have three storage targets: two global accessible file systems (\texttt{lustre01} and \texttt{lustre02}), and one local file system in \texttt{/tmp} that can be accessed via the POSIX backend.
Each of them comes with a lightweight performance model and the maximum size of data fragments.
The metadata section (Line 24) utilises here a POSIX interface to store the information about the ESDM objects. Internally, ESDM creates so-called containers and dataset objects to manage data fragments.

@@

The user now has to provide information about datasets required for input and the generated output for each workflow task in a separate file similarly to Cylc's workflow configuration file.
An example of an I/O-workflow configuration file (\texttt{io.cylc}) is shown in \Cref{lst:cylc}. In this file, information about Task\,1 is given by example, and we expect the extra information about all tasks in the same file. Ultimately, this could be integrated into the workflow specification file of Cylc.

@@

From the list of opportunities, we realise data placement and migration in a heterogeneous (multi-storage) environment.
These goals will be achieved via the proposed I/O-aware scheduler, called here EIOS (ESDM I/O Scheduler).
EIOS will make the schedule considering Cylc workflow, and ESDM provided system characteristics.
Components of it are involved in different steps of the workflow and the I/O path.
While Cylc schedules the workflow, EIOS can provide hints about colocating tasks which provide the opportunity for keeping data in local storage.
Our design imposes only minor changes to Cylc as the core requirements are covered by normal functionalities:

@@

This will be achieved by defining environment variables in the job script which are set to the output of another external command (run on the job host).
This command, provided by EIOS, will have access to all the normal Cylc environment variables providing details about the current task.
\end{description}

@@

The steps to execute a workflow enriched with I/O information and perform smarter scheduling is depicted in \cref{pic/stages-io}.
The suggested alterations can be seen in boxes pointed by red arrows, and the remaining components are the current state-of-the-art for workflows in climate and weather from \cref{pic/stages}.
In the following, we describe the modifications we propose in this vision paper for each component involved in the software stack.

>>>>

The vision proposed here is built on the belief that workflows composed of data, computing, and communication-intensive tasks should drive interfaces and hardware configurations to best support the programming models.
When delivered, this work will increase the opportunity for smarter scheduling of computing by considering storage in heterogeneous storage systems.
We illustrate the performance-impact on an example workload using a model built on measured performance data using ESDM at DKRZ.

@@

\subsection{Data Centre Infrastructure}

@@

The abstraction and automation of the I/O inside a workflow allow a runtime system to improve data placement and apply data reduction on heterogeneous storage systems.
Taking into consideration the architecture and workflow information, a smarter schedule can now be realised by exploiting the additional information.
Value and priority can influence fault-tolerance strategies and imply the quality of service for performance and availability.
Aspects like data reproducibility (can it be recomputed easily), type of the experiment (test, production), and runtime constraints for the overall and potential workflow could allow reducing costs and, hence, increasing scientific output.
Next, we outline the two core strategies and potential the proposed vision can bring to improve current workflows.
In the design proposed in this work (\Cref{sec:design}), we will focus on the data placement strategy.

\paragraph{Strategy: Data Placement}

@@

\paragraph{Strategy: Data Reduction}

@@

ESDM is used as I/O middleware in the parallel application (with NetCDF or directly) and orchestrates the I/O according to a simplified ESDM configuration file (\texttt{esdm.conf}).
This file contains information about the available technology in the data centre, its I/O characteristics, and will be used to make decisions about how to prioritise I/O targets.
In the example presented in \cref{lst:esdm.conf}, we have three storage targets: two global accessible file systems (\texttt{lustre01} and \texttt{lustre02}), and one local file system in \texttt{/tmp} that can be accessed via the POSIX backend.
Each of them comes with a lightweight performance model and the maximum size of data fragments.
The metadata section (Line 24) utilises here a POSIX interface to store the information about the ESDM objects. Internally, ESDM creates so-called containers and dataset objects to manage data fragments.

@@

The user now has to provide information about datasets required for input and the generated output for each workflow task in a separate file similarly to Cylc's workflow configuration file.
An example of an I/O-workflow configuration file (\texttt{io.cylc}) is shown in \cref{lst:cylc}. In this file, information about Task\,1 is given by example, and we expect the extra information about all tasks in the same file. Ultimately, this could be integrated into the workflow specification file of Cylc.

@@

From the list of opportunities, we realise data placement and migration in a heterogeneous (multi-storage) environment.
These goals will be achieved via the proposed I/O-aware scheduler, called here EIOS (ESDM I/O Scheduler).
EIOS will make the schedule considering Cylc workflow and ESDM provided system characteristics.
We are working together with Cylc Team in developing how EIOS interfaces with Cylc.
% Components of it are involved in different steps of the workflow and the I/O path.
While Cylc schedules the workflow, EIOS can provide hints about colocating tasks which provide the opportunity for keeping data in local storage.
Our design imposes only minor changes to Cylc as the core requirements are covered by normal functionalities:

@@

This will be achieved by defining environment variables in the job script which are set to the output of another external command (run on the job host).
This command, also provided by EIOS, will have access to all the normal Cylc environment variables with details about the current task.
\end{description}

@@

The steps to execute a workflow enriched with I/O information and perform smarter scheduling is depicted in \cref{pic/stages-io}.
Components of EIOS are involved in different steps of the workflow and the I/O path.
The suggested alterations can be seen in boxes pointed by red arrows, and the remaining components are the current state-of-the-art for workflows in climate and weather from \cref{pic/stages}.
In the following, we describe the modifications we propose in this vision paper for each component involved in the software stack.
